Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2017-11-10 paper.tex
Add Matt's comments through p. 13

We don't show it here, but the angular distribution contributing to the
\footnote{TODO: do show it}
total number density is also strongly affected by elastic scattering.
Without elastic scattering, the central regions emitting 60\%
of the neutrinos for the species $\{\nu_e,\bar{\nu}_e,\nu_x\}$
have scales $R_\nu\approx\{40,35,20\}$~km;
with elastic scattering the scales are
$R_\nu\approx\{50,45,35\}$~km.
% From by-hand summation over histogram G(cosA_i)*dcosA
% Script: $DIR/plot-compare_scat_vs_noscat-G_angular_distribution-all.gnu
% DIR=.../20161007-ray_tracing_tests/CollapseProfile/ResultsBrett-B-03

@@

\subsection{Hypermassive Neutron Star Remnant:
  Testing Everything}
\label{ssec:test_disk_comparison}
The merger of two neutron stars by gravitational wave emission produces a
postmerger configuration composed of a single neutron star surrounded by a disk.
Because of strong differential rotation, the remnant temporarily avoids
collapse to a black hole, even if its mass exceeds the threshold of dynamical
instability for a rigidly rotating neutron star \cite{duez2009-review}.
These objects, called hypermassive neutron stars,
may avoid collapse for $X\textrm{--}Y$~s,
\footnote{TODO: cite $X$, $Y$}
depending upon a number of physical factors including thermal pressure,
magnetic fields, and the microphysics of the fluid.

@@

We use a single time snapshot from that configuration
at $t=11\,{\rm ms}$ after merger.
The volume data are represented on a pseudospectral domain of a
sphere with concentric shells extending to $r\approx1400$~km.
Radial grid spacings are $\Delta r \approx 0.15$~km in the star
and $\Delta r \approx 2.5$~km in the disk,
with 12 cells spanning polar angles $\theta\in[0,\pi]$
and 24 cells spanning azimuthal angles $\phi\in[0,2\pi]$.
\footnote{TODO: way too-low angular resolution}
Though a pseudospectral representation of non-smooth hydrodynamic data
introduces some Gibbs-like oscillations in the variables,
we choose to use this representation instead of the mesh-refined
finite difference grid of the evolution because the pseudospectral
representation uses much less memory:
95~MB in pseuspectral vs.\ 3~GB in finite difference representation.
Obviously, memory loads of this order are not insurmountable;
but they would require some modifications to our
volume data interpolation infrastructure.
For the purposes of this analysis the pseudospectral representation is adequate.
Figs.~\ref{fig:nsns_rho_merid}--\ref{fig:nsns_temp_equat}
show slices of density and temperature from the finite difference volume data.

@@

The distributions of radial fluxes over observer position are shown
in Figs.~\ref{fig:nsns_theta_distrib_Kr} and \ref{fig:nsns_theta_distrib_Hr}.
With the \emph{noscat} treatment the $\nu_x$ luminosities are relatively
constant in $\theta$, because the disk is optically thin to $\nu_x$,
and most of the neutrinos come from the hypermassive neutron star,
which is roughly spherical.
The slight upward trend in $\nu_x$ luminosities with $\theta$ may be due to
asymmetries in the configuration.
It is not due to the Doppler shift from the rapid rotation of the star
(a hypothesis we tested by setting $u_i=0$ and $W=1$ in the ray
tracing equations).
Because of their larger absorption cross section,
the disk is not optically thin to $\nu_e$ and $\bar{\nu}_e$,
and observers within view of the hot hypermassive neutron star measure
the largest radial fluxes.

@@

Fig.~\ref{fig:nsns_theta_distrib_avg_eps} shows the distribution of
average energies of radial fluxes over observer positions.
With and without scattering, the average energies are highest
for observers near the polar axis, since polar observers get a
direct view of the hot hypermassive neutron star.
Scattering decreases average energies across all observer positions,
as it did in the post-bounce configuration, or any configuration of
a hot interior surrounded by a scattering envelope.
The strength of the effect of scattering on the average energies of
the different species is seen to follow the ranking
$\nu_x>\bar{\nu}_e>\nu_e$ due to two factors:
the average energies of the spectra obey the same ranking,
and the thicknesses of the different species' scattering envelopes
also obey the same ranking.
As we showed in the optically thin limit of
Eqn.~\ref{eqn:fse_homogeneous_thin_limit},
\footnote{TODO: too simplistic}
the scattering contribution $f_{\rm SE}$ is
proportional to energy $\varepsilon^2$ and path length $s$.

@@

\subsection{Neutrino Oscillations at High Neutrino Densities}
\label{ssec:V_nunu}
The Boltzmann Equation (Eqn.~\ref{eqn:boltzmann}) is one limiting
form of the quantum kinetic equations \cite{vlas2014-neutrino_qkes},
the limit where collisional mean free paths are much shorter
than oscillation lengths (i.e.\ $\varepsilon/\mathscr{K} \ll |H|^{-1}$
with $H$ the Hamiltonian describing coherent forward scattering interactions).
In this limit the neutrino density matrix takes the form
\begin{equation}
  \label{eqn:density_rho_0}
  \rho_0(\varepsilon,\ell_\mu) =
  \sum\limits_{\alpha}f^{\nu_\alpha}(\varepsilon,\ell_\mu)
  \ket{\nu_\alpha} \bra{\nu_\alpha},
\end{equation}
with $\ket{\nu_\alpha}$ the neutrino flavor eigenstates.
The neutrino state remains pure, or diagonal in the flavor basis.

>>>>

Although not shown here, the angular distribution contributing to the
total number density is also strongly affected by elastic scattering.
Without elastic scattering, the central regions emitting 60\%
of the neutrinos for the species $\{\nu_e,\bar{\nu}_e,\nu_x\}$
have scales $R_\nu\approx\{40,35,20\}$~km;
with elastic scattering the scales are
$R_\nu\approx\{50,45,35\}$~km.
% From by-hand summation over histogram G(cosA_i)*dcosA
% Script: $DIR/plot-compare_scat_vs_noscat-G_angular_distribution-all.gnu
% DIR=.../20161007-ray_tracing_tests/CollapseProfile/ResultsBrett-B-03

@@

\subsection{Hypermassive Neutron Star Remnant:
  Testing Everything}
\label{ssec:test_disk_comparison}
The merger of two neutron stars by gravitational wave emission produces a
postmerger configuration composed of a single neutron star surrounded by a disk.
Because of strong differential rotation and shock heating,
the remnant temporarily avoids
collapse to a black hole, even if its mass exceeds the threshold of dynamical
instability for a rigidly rotating neutron star \cite{duez2009-review}.
These objects, called hypermassive neutron stars,
may avoid collapse for $X\textrm{--}Y$~s,
\footnote{TODO: cite $X$, $Y$}
depending upon a number of physical factors including thermal pressure,
magnetic fields, and the microphysics of the fluid.

@@

We use a single time snapshot from that configuration
at $t=11\,{\rm ms}$ after merger.
In this way we approximate the system as stationary over a light-crossing time
of around 1~ms, which is far below the thermal timescale of the remnant.
The volume data are represented on a pseudospectral domain of a
sphere with concentric shells extending to $r\approx1400$~km.
Radial grid spacings are $\Delta r \approx 0.15$~km in the star
and $\Delta r \approx 2.5$~km in the disk,
with 12 cells spanning polar angles $\theta\in[0,\pi]$
and 24 cells spanning azimuthal angles $\phi\in[0,2\pi]$.
\footnote{TODO: way too-low angular resolution}
Though a pseudospectral representation of non-smooth hydrodynamic data
introduces some Gibbs-like oscillations in the variables,
we choose to use this representation instead of the mesh-refined
finite difference grid of the evolution because the pseudospectral
representation uses much less memory:
95~MB in pseuspectral vs.\ 3~GB in finite difference representation.
Obviously, memory loads of this order are not insurmountable;
but they would require some modifications to our
volume data interpolation infrastructure.
For the purposes of this analysis the pseudospectral representation is adequate.
Figs.~\ref{fig:nsns_rho_merid}--\ref{fig:nsns_temp_equat}
show slices of density and temperature from the finite difference volume data.

@@

The distributions of radial fluxes over observer position are shown
in Figs.~\ref{fig:nsns_theta_distrib_Kr} and \ref{fig:nsns_theta_distrib_Hr}.
With the \emph{noscat} treatment the $\nu_x$ luminosities are relatively
constant in $\theta$, because the disk is optically thin to $\nu_x$,
and most of the neutrinos come from the hypermassive neutron star,
which is roughly spherical.
The slight upward trend in $\nu_x$ luminosities with $\theta$ may be due to
asymmetries in the configuration.
It is not due to the Doppler shift or relativistic beaming
from the rapid rotation of the star,
a hypothesis we tested by setting $u_i=0$ and $W=1$ in the ray
tracing equations.
Because of their larger absorption cross section,
the disk is not optically thin to $\nu_e$ and $\bar{\nu}_e$,
and observers within view of the hot hypermassive neutron star measure
the largest radial fluxes of these species.

@@

Fig.~\ref{fig:nsns_theta_distrib_avg_eps} shows the distribution of
average energies of radial fluxes over observer positions.
With and without scattering, the average energies are highest
for observers near the polar axis, since polar observers get a
direct view of the hot hypermassive neutron star.
Scattering decreases average energies across all observer positions,
as it did in the post-bounce configuration, or any configuration of
a hot interior surrounded by a scattering envelope.
The strength of the effect of scattering on the average energies of
the different species is seen to follow the ranking
$\nu_x>\bar{\nu}_e>\nu_e$ due to two factors:
the average energies of the spectra obey the same ranking,
and the thicknesses of the different species' scattering envelopes
also obey the same ranking.
As we showed in the toy models with a homogeneous background scattering field
the scattering contribution $f_{\rm SE}$ is
proportional to energy $\varepsilon^2$ and path length $s$
(Eqn.~\ref{eqn:fse_homogeneous_thin_limit}).

@@

\subsection{Neutrino Oscillations at High Neutrino Densities}
\label{ssec:V_nunu}
The Boltzmann Equation (Eqn.~\ref{eqn:boltzmann}) is one limiting
form of the quantum kinetic equations for neutrinos
\cite{vlas2014-neutrino_qkes},
the limit where collisional mean free paths are much shorter
than oscillation lengths, i.e.\ $\varepsilon/\mathscr{K} \ll |H|^{-1}$
with $H$ the Hamiltonian describing coherent forward scattering interactions.
In this limit the neutrino density matrix takes the form
\begin{equation}
  \label{eqn:density_rho_0}
  \rho_0(\varepsilon,\ell_\mu) =
  \sum\limits_{\alpha}f^{\nu_\alpha}(\varepsilon,\ell_\mu)
  \ket{\nu_\alpha} \bra{\nu_\alpha},
\end{equation}
with $\ket{\nu_\alpha}$ the neutrino flavor eigenstates.
The neutrino state remains pure, or diagonal in the flavor basis.
